---
title: "Getting Started with genepicoloc"
format: html
vignette: >
  %\VignetteIndexEntry{Getting Started with genepicoloc}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**genepicoloc** provides a scalable pipeline for genetic colocalization analysis
across molecular and clinical trait datasets. It includes a standalone
implementation of the approximate Bayes factor (ABF) colocalization method,
preprocessing tools (liftOver, variant naming), and integration with public
resources (eQTL Catalogue, GWAS Catalog).

The typical workflow:

1. **Preprocess** your GWAS summary statistics (liftOver, variant naming)
2. **Identify** significant genomic regions
3. **Define** target datasets (eQTL, pQTL, PheWAS)
4. **Run** colocalization across all region-dataset combinations
5. **Annotate** and explore results

### Example: CKDGen eGFR Kidney Atlas

Throughout this vignette, we use the CKDGen round 4 eGFR GWAS
(Wuttke et al. 2019, PMID: 31152163) as a running example. This analysis
identified 259 genome-wide significant loci and tested colocalization against
11 molecular and clinical datasets (~16,000 phenotypes), producing 6,439
high-confidence colocalizations.

## Installation

```{r}
# Install from GitHub
remotes::install_github("genepi-freiburg/genepicoloc")
```

**System requirements:** `bgzip`, `tabix` (from htslib), and `tar` must be
available on your `PATH`.

For preprocessing (optional): UCSC `liftOver` binary and chain file.

## Step 1: Preprocess Summary Statistics

### If your GWAS is already in GRCh38

Load and format directly:

```{r}
library(genepicoloc)
library(data.table)

sumstats_1 <- fread("my_gwas_hg38.txt.gz")

# Rename to standard columns
sumstats_1 <- match_cols(
  sumstats = sumstats_1,
  Name = "variant_id",    # chr:pos:ref:alt format
  rsID = "rsid",          # optional, can be NA
  CHR  = "chromosome",
  POS  = "position",
  A1   = "effect_allele",
  A2   = "other_allele",
  BETA = "beta",
  SE   = "se",
  nlog10P = "nlog10p",    # -log10(p-value)
  AF   = "eaf",           # needed for quantitative traits
  N    = "n"              # needed for quantitative traits
)
```

### If your GWAS is in hg19

Most published GWAS use GRCh37/hg19 coordinates. Use `genepi_liftover()` to
convert:

```{r}
# CKDGen eGFR example (hg19 -> hg38)
sumstats_raw <- fread("20171016_MW_eGFR_overall_ALL_nstud61.dbgap.txt.gz")

sumstats_hg38 <- genepi_liftover(
  sumstats = sumstats_raw,
  CHR_name = "Chr",
  POS_name = "Pos_b37",
  A1_name  = "Allele1",
  A2_name  = "Allele2",
  liftOver_bin   = "liftOver",
  liftOver_chain = "hg19ToHg38.over.chain.gz"
)

# IMPORTANT: genepi_liftover() returns a data.frame
# Convert back if you need data.table syntax
setDT(sumstats_hg38)
```

The function adds `CHR_hg38`, `POS_hg38`, `A1_hg38`, `A2_hg38` columns.

**Where to get liftOver:**

- Binary: <https://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/liftOver>
- Chain: <https://hgdownload.soe.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz>

### Add variant names and rsIDs

To standardize variant naming (chr:pos:ref:alt) and add rsIDs:

```{r}
sumstats_named <- name_by_position(
  sumstats = sumstats_hg38,
  CHR_name = "CHR_hg38",
  POS_name = "POS_hg38",
  A1_name  = "A1_hg38",
  A2_name  = "A2_hg38",
  dbSNP_dir = "/path/to/ensembl_vcf/"
)
```

Per-chromosome Ensembl VCF files can be downloaded from:
<https://ftp.ensembl.org/pub/release-115/variation/vcf/homo_sapiens/>

### Calculate -log10(P)

```{r}
# Standard case
sumstats_hg38$nlog10P <- -log10(sumstats_hg38[["P-value"]])

# For extreme p-values (e.g., p < 1e-324)
sumstats_hg38$nlog10P <- handle_underflow(
  sumstats_hg38[["P-value"]],
  return_nlog10P = TRUE
)
```

### Required columns

| Column | Description | Required |
|--------|-------------|----------|
| `Name` | chr:pos:ref:alt (e.g., chr16:20350000:A:G) | yes |
| `rsID` | dbSNP identifier | optional (can be NA) |
| `CHR` | Chromosome (1-22, X) | yes |
| `POS` | Position (GRCh38) | yes |
| `A1`, `A2` | Effect and other allele | yes |
| `BETA`, `SE` | Effect size and standard error | yes |
| `nlog10P` | -log10(p-value) | yes |
| `AF` | Allele frequency | quantitative traits only* |
| `N` | Sample size | quantitative traits only* |

*AF and N are needed when `sdY` is not provided (the usual case).
For case-control traits, only BETA and SE are required.

## Step 2: Identify Significant Regions

```{r}
coloc_regions_list <- get_coloc_regions(
  sumstats = sumstats_1,
  CHR_name = "CHR",
  POS_name = "POS",
  nlog10p_value_name = "nlog10P",
  nlogP_threshold = 7.30103,   # p = 5e-8
  halfwindow = 500000           # 500 kb each side
)

# CKDGen eGFR: 259 significant regions across 22 chromosomes
nrow(coloc_regions_list$coloc_regions_PASS)
```

Write regions to indexed files for fast retrieval:

```{r}
sumstats_1_file <- write_regions(
  coloc_regions_list,
  sumstats_name = "output/eGFR"
)
```

## Step 3: Define Target Datasets

### Option A: YAML configuration (recommended for custom analyses)

```yaml
# studies.yaml
eQTLGen:
  description: "eQTLGen blood eQTL"
  query_function: retrieve_sumstats_tabix
  type: quant
  sdY: 1
  files:
    - /path/to/eQTLGen/cis_eQTLs_hg38.tsv.gz

UKB_PPP:
  description: "UK Biobank plasma proteomics"
  query_function: retrieve_sumstats_tabix
  type: quant
  sdY: 1
  path: /path/to/UKB_PPP/
  pattern: "*.txt.gz"
```

```{r}
args_df <- create_args_df_from_config("studies.yaml")
```

### Option B: eQTL Catalogue API

```{r}
# Browse available datasets
datasets <- get_datasets_eQTL_Catalogue()

# Create args for GTEx kidney cortex
eqtl_args <- make_eQTL_Catalogue_args(
  QTS = "QTS000015",
  quant_method = "ge"
)
```

## Step 4: Run Colocalization

```{r}
sumstats_1_args <- list(
  coloc_regions_PASS = coloc_regions_list$coloc_regions_PASS,
  sumstats_1_function = "retrieve_sumstats_tabix",
  sumstats_1_file = sumstats_1_file,
  sumstats_1_type = "quant",   # or "cc" for case-control
  sumstats_1_sdY = NA
)

genepicoloc_wrapper(
  dir_out = "output/eGFR_coloc",
  sumstats_1_args = sumstats_1_args,
  args_df = args_df,
  mc_cores = 10
)
```

The CKDGen eGFR atlas ran 259 regions x 15,930 datasets = 4.1 million
colocalization tests in ~17 hours on 20 cores.

### Saving regional summary statistics

By default, only colocalization results (posterior probabilities) are saved.
To also save the underlying summary statistics for each region-dataset pair
(needed for regional association plots), use `save_sumstats = TRUE`:

```{r}
genepicoloc_wrapper(
  ...,
  save_sumstats = TRUE,
  force_save_sumstats = TRUE  # skip interactive confirmation
)
```

**Warning:** This generates substantial disk usage (tens of GB for large runs).
Use only when you need regional plots or downstream fine-mapping.

### Standalone colocalization

For simple pairwise tests without the full pipeline:

```{r}
# Prepare datasets
d1 <- list(
  beta = sumstats_1$BETA,
  varbeta = sumstats_1$SE^2,
  snp = sumstats_1$Name,
  type = "quant",
  sdY = 1
)

d2 <- list(
  beta = sumstats_2$BETA,
  varbeta = sumstats_2$SE^2,
  snp = sumstats_2$Name,
  type = "quant",
  sdY = 1
)

# Run colocalization (no external dependencies)
result <- coloc_abf(d1, d2)

# Posterior probabilities
result$summary
#>     nsnps PP.H0.abf PP.H1.abf PP.H2.abf PP.H3.abf PP.H4.abf
#>      1500     0.000     0.000     0.000     0.015     0.985
```

## Step 5: Annotate Results

After the pipeline completes:

```{r}
# Results are automatically consolidated into per-study files
# in output/eGFR_coloc/unfilt/

# Load combined filtered results
results <- readRDS("output/eGFR_coloc/annot/annot_filt.RDS")
```

### CKDGen eGFR Atlas Results

The eGFR analysis produced 6,439 high-confidence colocalizations (PP.H4 > 0.8)
across 11 studies:

| Study | Colocalizations | Description |
|-------|-----------------|-------------|
| MVP_R4_EUR | 2,504 | PheWAS (creatinine, eGFR, CKD, hypertension) |
| UKB_PPP_EUR | 1,639 | Plasma proteomics (2,940 proteins) |
| Icelanders_pGWAS | 787 | Plasma proteomics (4,907 proteins) |
| Kidney_eQTL | 498 | Kidney-specific gene expression |
| CKDGen_r4 | 333 | Other kidney traits (BUN, urate, UACR) |
| FinnGen_r9 | 281 | Clinical endpoints (2,272 diseases) |
| GCKD_mGWAS_urine | 103 | Urine metabolomics |
| eQTLGen | 89 | Blood eQTL |
| GCKD_mGWAS_plasma | 70 | Plasma metabolomics |
| UKB_TOPMed | 69 | UK Biobank clinical phenotypes |
| UKB_kidney_vol | 66 | Kidney MRI volumes |

Top regions by number of colocalizing studies (out of 11):

| Region | Gene | Studies | Colocalizations |
|--------|------|---------|-----------------|
| chr12 | ACAD10 | 9 | 745 |
| chr2 | GCKR | 9 | 727 |
| chr16 | PDILT/UMOD | 8 | 292 |
| chr6 | HLA | 8 | 172 |
| chr2 | CPS1 | 8 | 140 |
| chr4 | SHROOM3 | 6 | 105 |
| chr4 | FGF5 | 6 | 81 |

## Key Concepts

### Hypotheses

Colocalization tests 5 hypotheses:

- **H0**: No association with either trait
- **H1**: Association with trait 1 only
- **H2**: Association with trait 2 only
- **H3**: Both associated, different causal variants
- **H4**: Both associated, **shared causal variant** (colocalization)

**PP.H4 > 0.8** is the standard threshold for evidence of colocalization.

### sumstats_1 vs sumstats_2

- **sumstats_1** (primary): your GWAS of interest - the "query"
- **sumstats_2** (secondary): datasets to colocalize against - the "targets"

## Session Info

```{r, eval=TRUE}
sessionInfo()
```
