---
title: "Getting Started with genepicoloc"
format: html
vignette: >
  %\VignetteIndexEntry{Getting Started with genepicoloc}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**genepicoloc** provides a scalable pipeline for genetic colocalization analysis
across multiple QTL and PheWAS datasets. It wraps the
[coloc](https://chr1swallace.github.io/coloc/) package with infrastructure for
handling thousands of dataset comparisons efficiently.

The typical workflow has five steps:

1. **Format** your primary GWAS summary statistics (sumstats_1)
2. **Identify** significant genomic regions
3. **Define** secondary datasets (sumstats_2) via YAML configuration
4. **Run** colocalization across all region-dataset combinations
5. **Consolidate** and annotate results

## Installation

```{r}
# Install from GitHub
remotes::install_github("genepi-freiburg/genepicoloc")
```

**System requirements:** `bgzip`, `tabix` (from htslib), and `tar` must be
available on your `PATH`.

## Step 1: Load and Format Primary Summary Statistics

Your primary GWAS (sumstats_1) needs standard column names. Use `match_cols()`
to rename columns to the expected format.

```{r}
library(genepicoloc)
library(data.table)

# Load your GWAS summary statistics
sumstats_1 <- fread("my_gwas.txt.gz")

# Rename columns to standard format
sumstats_1 <- match_cols(
  sumstats = sumstats_1,
  Name = "variant_id",    # chr:pos:ref:alt format
  rsID = "rsid",
  CHR  = "chromosome",
  POS  = "position",
  A1   = "effect_allele",
  A2   = "other_allele",
  BETA = "beta",
  SE   = "se",
  P    = "pvalue",
  AF   = "eaf",
  N    = "n"
)

# Calculate -log10(P) with underflow handling
sumstats_1[, nlog10P := handle_underflow(P, return_nlog10P = TRUE)]

# Basic QC
sumstats_1 <- sumstats_1[is.finite(nlog10P) & !is.na(BETA) & !is.na(SE) & SE > 0]
```

### Required columns

| Column | Description |
|--------|-------------|
| `Name` | Variant ID in chr:pos:ref:alt format (e.g., `chr16:17906244:T:C`) |
| `CHR` | Chromosome (1-22, X) |
| `POS` | Genomic position (GRCh38) |
| `A1` | Effect allele |
| `A2` | Other allele |
| `BETA` | Effect size |
| `SE` | Standard error |
| `nlog10P` | -log10(p-value) |
| `AF` | Allele frequency |
| `N` | Sample size |

### Preprocessing: LiftOver and rsID annotation

If your GWAS is on hg19, use `genepi_liftover()` to convert coordinates:

```{r}
sumstats_hg38 <- genepi_liftover(
  sumstats = sumstats_1,
  CHR_name = "CHR", POS_name = "POS",
  A1_name = "A1", A2_name = "A2",
  liftOver_bin = "liftOver",
  liftOver_chain = "hg19ToHg38.over.chain.gz"
)
```

To add rsIDs from Ensembl VCF files:

```{r}
sumstats_named <- name_by_position(
  sumstats = sumstats_hg38,
  CHR_name = "CHR_hg38", POS_name = "POS_hg38",
  A1_name = "A1_hg38", A2_name = "A2_hg38",
  dbSNP_dir = "/path/to/ensembl_vcf/"
)
```

## Step 2: Identify Significant Regions

`get_coloc_regions()` finds genome-wide significant loci and defines windows
around them:

```{r}
coloc_regions_list <- get_coloc_regions(
  sumstats = sumstats_1,
  CHR_name = "CHR",
  POS_name = "POS",
  nlog10p_value_name = "nlog10P",
  nlogP_threshold = 7.30103,   # p = 5e-8
  halfwindow = 500000           # 500 kb each side
)

# Check how many regions were found
coloc_regions_list$coloc_regions_PASS
```

Then write the regions to indexed files for fast retrieval:

```{r}
sumstats_1_subset_file <- write_regions(
  coloc_regions_list,
  sumstats_name = "output/my_gwas"
)
```

## Step 3: Define Secondary Datasets via YAML

Create a YAML configuration file listing your target datasets. This is the
recommended way to define sumstats_2 sources:

```yaml
# studies.yaml

eQTLGen:
  description: "eQTLGen blood eQTL (Vosa et al. 2021)"
  query_function: retrieve_sumstats_tabix
  type: quant
  sdY: 1
  files:
    - /path/to/eQTLGen/cis_full_chr1.txt.gz
    - /path/to/eQTLGen/cis_full_chr2.txt.gz

GTEx_eQTL:
  description: "GTEx v8 eQTL (all tissues)"
  query_function: retrieve_sumstats_tabix
  type: quant
  sdY: 1
  path: /path/to/GTEx/
  pattern: "*.allpairs.txt.gz"
  expected_count: 49

FinnGen:
  description: "FinnGen release 9"
  query_function: retrieve_sumstats_tabix
  type: cc
  path: /path/to/FinnGen/
  pattern: "*.gz"
```

Three file source modes are supported:

- **`files:`** - explicit list of file paths
- **`path:` + `pattern:`** - directory listing with glob pattern
- **`metadata_file:`** - TSV with per-file metadata (columns: `file`, `type`, `sdY`)

Browse and validate your config:

```{r}
# List available studies
list_available_studies("studies.yaml")

# Build the args_df for the pipeline
args_df <- create_args_df_from_config("studies.yaml")

# Or select specific studies
args_df <- create_args_df_from_config(
  "studies.yaml",
  selected_studies = c("eQTLGen", "GTEx_eQTL")
)
```

See `system.file("extdata", "example_studies.yaml", package = "genepicoloc")`
for more examples.

## Step 4: Run Colocalization

Assemble the sumstats_1 arguments and run:

```{r}
sumstats_1_args <- list(
  coloc_regions_PASS = coloc_regions_list$coloc_regions_PASS,
  sumstats_1_function = "retrieve_sumstats_tabix",
  sumstats_1_file = sumstats_1_subset_file,
  sumstats_1_type = "quant",   # or "cc" for case-control
  sumstats_1_sdY = NA           # NA for case-control, or provide sdY
)

genepicoloc_wrapper(
  dir_out = "output/coloc_results",
  sumstats_1_args = sumstats_1_args,
  args_df = args_df,
  mc_cores = 4,
  save_sumstats = FALSE  # set TRUE to keep sumstats for regional plots
)
```

Key parameters:

- `mc_cores` - number of parallel cores
- `save_sumstats` - save filtered sumstats for downstream use (regional plots,
  fine-mapping). Increases disk usage substantially.
- `max_regions_per_job` - regions per batch (default 10; increase for single-phenotype studies)
- `debug_mode` - run sequentially for troubleshooting

## Step 5: Consolidate and Annotate Results

After the pipeline completes, consolidate results from all jobs:

```{r}
results <- consolidate_coloc_results(
  dir_out = "output/coloc_results",
  verbose = TRUE
)
```

Add gene annotations:

```{r}
results <- add_gene_annotations(results, verbose = TRUE)
```

The output data.table contains one row per region-dataset pair with coloc
posterior probabilities:

| Column | Description |
|--------|-------------|
| `PP.H0.abf` | P(no association in either) |
| `PP.H1.abf` | P(association in sumstats_1 only) |
| `PP.H2.abf` | P(association in sumstats_2 only) |
| `PP.H3.abf` | P(both associated, different causal variants) |
| `PP.H4.abf` | P(both associated, shared causal variant) |
| `nsnps` | Number of shared SNPs in the region |

**PP.H4 > 0.8** is commonly used as a threshold for evidence of colocalization.

## Data Sources

### tabix-indexed files

Most datasets use `retrieve_sumstats_tabix()` as the query function. Your
sumstats_2 files need to be bgzip-compressed and tabix-indexed:
```bash
bgzip my_sumstats.txt
tabix -s 1 -b 2 -e 2 -S 1 my_sumstats.txt.gz
```

The files must have standard columns (Name, CHR, POS, A1, A2, BETA, SE, AF, N)
or columns recognized by `format_sumstats()`.

### eQTL Catalogue

genepicoloc has built-in support for the
[eQTL Catalogue](https://www.ebi.ac.uk/eqtl/) API:

```{r}
# Browse available datasets
datasets <- get_datasets_eQTL_Catalogue()

# Create args_df for a specific dataset
args_df_eqtl <- make_eQTL_Catalogue_args(
  QTS = "QTS000015",      # dataset ID
  quant_method = "ge"      # gene expression
)
```

### Custom query functions

For non-standard file formats, write a custom query function that accepts
`sumstats_file`, `CHR`, `BP_START`, `BP_STOP` parameters and returns a
data.table with standard columns. Set `query_function` in YAML to your
function's name.

## Output Structure

```
output/coloc_results/
  jobs/
    job_0001/
      coloc.tar         # colocalization results (RDS per dataset)
      sumstats.tar      # sumstats if save_sumstats=TRUE
      sumstats_1.RDS    # primary dataset for this job's regions
    job_0002/
      ...
  unfilt/
    coloc_results.RDS   # consolidated results (after consolidate)
  metadata files
```

## Sumstats Concepts

- **sumstats_1** (primary): Your GWAS of interest - the "query" study
- **sumstats_2** (secondary): External datasets to colocalize against - the
  "targets" (eQTL, pQTL, PheWAS, other GWAS)

The pipeline tests each sumstats_2 dataset against sumstats_1 in every
significant region, producing PP.H4 posterior probabilities for shared causal
variants.

## Session Info

```{r, eval=TRUE}
sessionInfo()
```
